## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement, a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.

This analysis makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

Dataset: Activity monitoring data [activity.csv] in the GitHub repository
The variables included in this dataset are:
* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data
```{r cache=TRUE}
setwd("~/Personal-Folder/Study-Git/reprod-research-wk1a/RepData_PeerAssessment1")
activity_df <- read.csv("activity.csv", header=T, sep=",")
```

## What is mean total number of steps taken per day?
```{r cache=TRUE}
steps_per_day_df <- aggregate(activity_df$steps, by=list(activity_df$date), FUN=sum)
names(steps_per_day_df) <- c("date","steps")
hist(steps_per_day_df$steps, 
     main="Number of steps per day",
     breaks=10, 
     col="orange", 
     xlab="Steps")
mean_steps_per_day <- mean(steps_per_day_df$steps, na.rm=T)
median_steps_per_day <- median(steps_per_day_df$steps, na.rm=T)
```

The mean number of steps per day is  : `r as.character(mean_steps_per_day)` 

The median number of steps per day is: `r as.character(median_steps_per_day)`

## What is the average daily activity pattern?
```{r cache=TRUE}
mean_steps_per_5min_df <- aggregate(activity_df$steps, by=list(activity_df$interval), FUN=mean, na.rm=T)
names(mean_steps_per_5min_df) <- c("interval","mean_steps")
plot(mean_steps_per_5min_df$interval,
     mean_steps_per_5min_df$mean_steps, 
     ylab="Steps", 
     type="l", 
     xlab="Interval",
     main="Average per 5 minute interval",
     col="orange") 
interval_with_max_steps <- mean_steps_per_5min_df[mean_steps_per_5min_df$mean_steps==max(mean_steps_per_5min_df$mean_steps),1]
```

The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps is: `r interval_with_max_steps` 

## Imputing missing values
```{r cache=TRUE}
num_rows_missing_vals <- nrow(activity_df) - sum(complete.cases(activity_df))
```
There are `r num_rows_missing_vals` missing values in the dataset. 

Missing values are substituted with the interval averages calculated accross all 61 days. 
```{r cache=TRUE} 
# there are 288 intervals per day and 61 full days, merge the dataframes into a new set 
activity_df_clean <- merge(activity_df,mean_steps_per_5min_df,by.x="interval",by.y="interval",all=FALSE)
activity_df_clean$steps <- ifelse(is.na(activity_df_clean$steps),
                           activity_df_clean$mean_steps,activity_df_clean$steps)
steps_per_day_df_clean <- aggregate(activity_df_clean$steps, by=list(activity_df_clean$date), FUN=sum)
names(steps_per_day_df_clean) <- c("date","steps")
hist(steps_per_day_df_clean$steps, 
     main="Number of steps per day - Cleaned Data",
     breaks=10, 
     col="orange", 
     xlab="Steps")
mean_steps_per_day_clean <- mean(steps_per_day_df_clean$steps)
median_steps_per_day_clean <- median(steps_per_day_df_clean$steps)
```

The mean number of steps per day is  : `r as.character(mean_steps_per_day_clean)` 

The median number of steps per day is: `r as.character(median_steps_per_day_clean)`

## Are there differences in activity patterns between weekdays and weekends?
```{r cache=TRUE}
activity_df_clean$date <- strptime(activity_df_clean$date, "%Y-%m-%d")
activity_df_clean$weekend_or_day <- as.factor(ifelse(weekdays(activity_df_clean$date) %in% 
                                    c("Saturday","Sunday"),"weekend","weekday"))
mean_steps_per_5min_wde_df <- aggregate(activity_df_clean$steps,
                              by=list(activity_df_clean$weekend_or_day,activity_df_clean$interval), 
                              FUN=mean, 
                              na.rm=T)
names(mean_steps_per_5min_wde_df) <- c("weekend_or_day","interval","mean_steps")
# use the ggplot2 plotting system
library(ggplot2)
ggplot(mean_steps_per_5min_wde_df, 
       aes(interval, mean_steps)) +	   
	   geom_line(colour="orange") + 
	   facet_grid(weekend_or_day ~ .) + 
	   labs(x="5-minute interval") + 
	   labs(y="Number of steps")
```	   



